---
title: "STATS786 Group Project"
author: "Group 4: Joshua Buchanan, Bea Cooke, Xinyi Kong, Marlon Chhour"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    toc_depth: 3
    number_sections: true
    df_print: paged
    code_folding: show
editor_options: 
  markdown: 
    wrap: 72
---

```{css, echo=FALSE}
.contributor {
  background-color: #2c3e50;
  color: #ecf0f1;
  padding: 4px 10px;
  border-radius: 4px;
  font-size: 0.8em;
  font-family: monospace;
  display: inline-block;
  margin-bottom: 12px;
}

.discussion {
  background-color: #f8f9fa;
  border-left: 4px solid #18bc9c;
  padding: 12px 16px;
  margin: 12px 0;
  border-radius: 0 4px 4px 0;
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo      = TRUE,
  warning   = FALSE,
  message   = FALSE,
  fig.align = "center",
  fig.width = 9,
  fig.height = 5
)
```

```{r libraries, message=FALSE, warning=FALSE}
library(tidyverse)
library(fpp3)
library(tsibble)
library(feasts)
library(fable)
library(lubridate)
library(here)
library(urca)
library(imputeTS)
library(patchwork)
library(magrittr)
```

------------------------------------------------------------------------

# Data Cleaning and Exploratory Data Analysis

## Data Description

::: contributor
Contributors: Joshua Buchanan
:::

::: discussion
The dataset contains 993 observations of monthly average temperature
(°C) recorded at a single location, spanning from April 1943 to
approximately mid-2025: a period of over 80 years. Each row represents
one month, with the month name and year stored as separate columns.
:::

```{r load-data}
# Load data
nelson <- read_csv('nelson.csv') 
glimpse(nelson)
```

------------------------------------------------------------------------

## Train / Test Split

::: contributor
Contributors: Joshua Buchanan
:::

::: discussion
The training set excludes the most recent 12 months, which form the test
set. All model fitting is performed on the training set only.
:::

```{r train-test-split}
# Turn into tsibble first
nelson <- nelson %>%
  mutate(date = yearmonth(paste(Month, Year))) %>%
  select(date, Temperature) %>%
  as_tsibble(index = date)
glimpse(nelson)

# Create training and test sets: test is last 12 months.
train <- nelson %>% filter(date < max(date) - 11)
test <- nelson %>% filter(date >= max(date) - 11)

# Checking number of rows
nrow(train)
nrow(test)

# Checking the range too. 
range(train$date)
range(test$date)
```

------------------------------------------------------------------------

## Missing Values

::: contributor
Contributors: Joshua Buchanan
:::

::: discussion
The Nelson temperature series contains four missing values, all located
within the training set (the test set is complete). Rather than dropping
or mean-imputing these observations, we used ARIMA-based interpolation
as outlined in FPP3 §13.9. This fits an ARIMA model to the observed
training data and fills gaps with the conditional expectation given the
surrounding values, preserving the local temporal structure better than
simpler alternatives.

`fill_gaps()` first makes the implicit missing timestamps explicit, then
`interpolate()` replaces them with model-fitted values. The filled dates
were verified against the original gap locations to confirm all four
were successfully imputed before proceeding with modelling.
:::

```{r missing-values}
has_gaps(nelson)  # check for gaps in time index
gaps <- scan_gaps(nelson) # shows where the gaps are
gaps 

# There are no gaps in the Test set so we only need to update the training set. 
# interpolation with ARIMA as per fpp3 sec:13.9
nel_miss <- train %>% fill_gaps()
train_filled <- nel_miss %>% 
                model(ARIMA(Temperature)) %>% 
                interpolate(nel_miss)

has_gaps(train_filled) 
vals <- train_filled %>% filter(date %in% gaps$date)
vals
```

```{r, fig.cap="Time plot of Nelson temperatures with interpolated temperature values for missing dates."}
train_filled %>% autoplot(Temperature) + geom_point(data = vals, aes(x = date, y = Temperature, label = Temperature), inherit.aes = FALSE, colour = "red", alpha = 0.5, size = 2) + theme_bw()
```

```{r, fig.cap="Subseries time plot of Nelson temperatures until 1947, temperature value interpolated at March 1946, denoted as a red vertical line."}
train_filled %>% 
  filter(date < yearmonth("1947 Jan")) %>% 
  autoplot(Temperature) + 
    geom_vline(xintercept = yearmonth("1946 Mar"), 
               colour = "red")
```

------------------------------------------------------------------------

## Visualisation and Feature Discussion

::: contributor
Contributors: Joshua Buchanan
:::

::: discussion
The time series displays strong, consistent seasonality with
temperatures peaking in January (summer, \~17–18°C) and reaching a
trough in July (winter, \~6–7°C). The seasonal subseries plot confirms
this pattern is highly stable across all 80 years, with monthly means
well-defined and little variation in the seasonal shape over time. The
STL decomposition confirms the seasonality is additive, with the
seasonal component oscillating consistently between approximately −6°C
and +3°C throughout the entire series, showing no evidence of changing
amplitude over time.

The trend component from the STL decomposition reveals a positive trend,
rising from approximately 11°C in the early 1940s to around 14°C by the
mid-2020s. The trend is not monotonically increasing, with a notable dip
around the 1970s–1980s before accelerating upward from the 1990s onward.
The remainder component appears to behave like white noise, fluctuating
randomly within approximately ±2°C with no obvious structure, large
outliers, or evidence of heteroscedasticity, which is encouraging for
modelling.

The ACF confirms strong positive autocorrelation at lags of 12, 24, and
36 months and strong negative autocorrelation at lags of 6, 18, and 30
months, consistent with the annual seasonal cycle, while the PACF shows
significant spikes across the first 12 lags suggesting a complex
short-term autocorrelation structure that will need to be accounted for
in both the ARIMA and dynamic regression models.
:::

```{r time-plot}
# Time plot
train_filled %>% autoplot(Temperature)
```

```{r subseries-plot}
train_filled %>% gg_subseries()
```

```{r}
# Seasonality plot 
train_filled %>% gg_season(Temperature)
```

```{r}
# ACF plot
train_filled %>% ACF(Temperature, lag_max=48) %>% autoplot()
```

```{r}
# PACF plot
train_filled %>% PACF(Temperature, lag.max=48) %>% autoplot()
```

```{r}
# STL decomp
train_filled %>%
  model(STL(Temperature ~ trend(window=21))) %>%
  components() %>%
  autoplot()
```

------------------------------------------------------------------------

# ETS Models

## Model Exploration and Shortlisting

::: contributor
Contributors: Xinyi Kong
:::

::: discussion
**Methodology:** 
Candidate ETS models were selected based on strong
annual seasonality, a slowly evolving trend, and approximately constant
variance over time. These features suggest that additive error and
additive seasonal structures are appropriate for this dataset. The
manually selected ETS models include: 

\## ETS model equations

### ETS(AAA):

$$
y_t = \ell_{t-1} + b_{t-1} + s_{t-m} + \epsilon_t
$$

### ETS(ANA):

$$
y_t = \ell_{t-1} + s_{t-m} + \epsilon_t
$$

### ETS(AAdA):

$$
y_t = \ell_{t-1} + \phi b_{t-1} + s_{t-m} + \epsilon_t
$$

### Multiplicative benchmark models:

### ETS(M,A,A):

$$
y_t = (\ell_{t-1} + b_{t-1} + s_{t-m})(1 + \epsilon_t)
$$

### ETS(M,Ad,A):

$$
y_t = (\ell_{t-1} + \phi b_{t-1} + s_{t-m})(1 + \epsilon_t)
$$ 

Given these features, additive error and additive seasonal structures
were considered the primary modelling choice. The initial shortlist
focused on ETS models with additive seasonality, including:

ETS(AAA): full additive specification with trend and seasonality
ETS(ANA): no-trend specification to test whether a trend component is
necessary ETS(AAdA): damped trend specification to improve long-term
forecasting stability

In addition, multiplicative specifications were included as benchmark
models to assess whether a proportional error and seasonal structure
could provide improved fit. These include:

ETS(M,A,A) ETS(M,Ad,M)

Although multiplicative models are typically suited to series with
variance increasing with the level of the series, they were included for
completeness and empirical comparison.

An automatic ETS procedure was also fitted to select an optimal model
based on AICc across the full ETS model space.

An automatic ETS model was also fitted using ETS(Temperature), which
selects the optimal combination of error, trend, and seasonal components
by minimising AICc over all possible ETS specifications.
:::

```{r ets-manual}
ETS_fit <- train_filled %>%
  fill_gaps() %>%
  mutate(Temperature = na_seadec(Temperature)) %>% #Bea note: i think this can be removed if using train_filled
  model(additive = ETS(Temperature ~ error("A") + trend("A") + season("A")),
        ANA = ETS(Temperature ~ error("A") + trend("N") + season("A")),
        AAdA = ETS(Temperature ~ error("A") + trend("Ad") + season("A")),
        multiplicative = ETS(Temperature ~ error("M") + trend("A") + season("M")),
        hw = ETS(Temperature ~ error("M") + trend("Ad") + season("M")))

# move this code to section 6 
# ETS_fc = ETS_fit %>% 
#   forecast(h = 12) 
# 
# ETS_fc %>%
#   autoplot(nelson)
```

```{r ets-auto}
# Automatically fitted ETS model
auto_fit = train_filled %>%
  fill_gaps() %>%
  mutate(Temperature = na_seadec(Temperature)) %>% #Bea note: same again, I think this can be removed if using train_filled
  model(ETS(Temperature))
report(auto_fit) 

## AAdA becomes the best now here too with the NAs filled via ARIMA in the training data.
```

```{r ets-comparison}
# Model comparison (e.g. AICc, BIC, accuracy)
accuracy(ETS_fit) %>%
  select(.model, RMSE, MAE, MAPE, MASE) #Bea Q: MSE AMSE and MAE reported in glance can we get rid of this or want to keep for others?

glance(ETS_fit) %>% arrange(AICc)

ETS_fit %>%
  select(additive) %>%
  gg_tsresiduals()

ETS_fit %>%
  select(ANA) %>%
  gg_tsresiduals()

ETS_fit %>%
  select(AAdA) %>%
  gg_tsresiduals() 

ETS_fit %>%
  augment() %>%
  features(.resid, ljung_box, lag = 24) #Bea Q: do we need this now - or just report both ljung and resid plots for selected model in part 5? If keep do we need to fix dof for each model?
```

------------------------------------------------------------------------

## Selected Model and Justification

::: contributor
Contributors: Xinyi Kong
:::

::: discussion
**Selected model:** [Note from Joshua - now the AAdA is outperforming
the ANA model in terms of AIC and AICc since we have switched to using
the training set with the filled values - what do you think of this?]

Based on AICc and forecast accuracy, the ETS(ANA) model is selected as
the preferred model. It achieves the lowest AICc and competitive error
measures, while maintaining a simpler structure without a trend
component.

Although Ljung-Box tests indicate remaining autocorrelation in the
residuals for all candidate models, this is common in long seasonal time
series such as temperature data and does not invalidate the model for
forecasting purposes.

Therefore, ETS(ANA) is selected as the final model due to its balance of
parsimony, interpretability, and predictive performance.
:::

```{r ets-selected}
# Selected ETS model
ETS_final <- train %>% # update to train_filled data set
  fill_gaps() %>%
  mutate(Temperature = na_seadec(Temperature)) %>%
  model(ANA = ETS(Temperature ~ error("A") + trend("N") + season("A"))) #[update to AAdA]
report(ETS_final)

# move to section 6 - 
# ETS_fc_final <- ETS_final %>%
#   forecast(h = 12)
# autoplot(ETS_fc_final, train)
```

------------------------------------------------------------------------

## Model Equations

::: contributor
Contributors: Xinyi Kong
:::

::: discussion
-   **Measurement equation:** $$
    y_t = \ell_{t-1} + s_{t-m} + \epsilon_t
    $$

-   **State equations:**

Level: $$
\ell_t = \ell_{t-1} + \alpha \epsilon_t
$$

Seasonal: $$
s_t = s_{t-m} + \gamma \epsilon_t
$$

-   **Parameter estimates:**

$$
\alpha = 0.1076, \quad \gamma = 0.0001008, \quad \sigma^2 = 0.7298
$$

[NOTE: we will have to change the above since top model has changed?]
:::

------------------------------------------------------------------------

# ARIMA Models

## Transformations and Differencing

::: contributor
Contributors: Bea Cooke
:::

::: discussion
**Transformations:** 

The time plot shows no visual evidence of
multiplicative seasonality, seasonal fluctuations appear approximately
constant and do not appear to depend on the level of the series.
Additionally, the selected Box-Cox parameter,  $\lambda = 0.92$, is
sufficiently close to $1$. Given no visual evidence and $\lambda \approx 1$, no transformation is used. 

**Differencing:** 

The series is non-stationary, showing both additive seasonality and increasing trend, this is supported by the KPSS test result which produced a p-val of 0.01 suggesting that we reject the null hypothesis (i.e. reject stationarity). Differencing is therefore required to induce stationarity prior to fitting ARIMA models. STL seasonal strength was reported as 0.97, suggesting one seasonal difference to be applied to the 12 month seasonal period. On visual inspection of the time plot the resulting seasonally differenced series appears stationary, resembling a white noise process. The KPSS test of the seasonally differenced series produced a p-val of 0.1, supporting stationarity of the seasonally differenced series. 
:::

```{r arima-transformations}
lambda <- train_filled %>% features(Temperature, features = guerrero) %>% pull(lambda_guerrero) #box-cox lambda: approx 1 -> no transform
cat("Box-Cox Lambda: ", lambda, sep = "  ")

strength <- train_filled %>% features(Temperature, feat_stl) %>% pull(seasonal_strength_year) #STL strength > 0.67 ->  seasonal difference 
cat("STL Seasonal Strength: ", strength, sep = "  ")

train_filled %>% features(Temperature, unitroot_nsdiffs) #check auto selection # seasonal diffs
train_filled %<>% mutate(diff_temp = Temperature %>% difference(12)) #seasonal diff
train_filled %>% features(diff_temp, unitroot_ndiffs) #check auto selection # diffs after applying seasonal
```

```{r, fig.cap="Time plot of the series with strong annual seasonality and weak positive trend."}
train_filled %>% autoplot(Temperature) + geom_smooth(se = F) + theme_bw()
```

```{r, fig.cap = "Time plot of the seasonally differenced series, the differenced series resembles a white noise process centered approximately around 0, with no obvious seasonality or trend."}
train_filled %>% autoplot(diff_temp) + theme_bw()
```


```{r arima-unit-root}
train_filled %>% features(Temperature, unitroot_kpss) #KPSS on not differenced , sig p-val -> not stationary
train_filled %>% features(diff_temp, unitroot_kpss) #KPSS on seasonally differenced, non-sig p-val -> stationary

```

------------------------------------------------------------------------

## Model Exploration and Shortlisting

::: contributor
Contributors: Bea Cooke
:::

::: discussion
**Methodology:**  

In all manually specified models the difference terms $d$ and $D$ are chosen to match the differencing outlined above. Since no first difference and only one seasonal difference was needed the non-seasonal term was chosen as $d=0$ and the seasonal term as $D=1$. 

Visual inspection of both the ACF and PACF plots of the seasonally differenced series allows assessment of candidate AR/MA parameters ($p,q, P,Q$). The PACF shows exponential decay in the seasonal lags, whereas the ACF shows a significant spike at lag 12 and no further significant spikes at any other seasonal lag, this suggests a seasonal MA(1) component and no seasonal AR component. Consequently, seasonal terms for all candidate models are $P=0$ and $Q=1$. 

The ACF shows further non-seasonal spikes at lags 1, 3 and 4, however spikes at lags 3 and 4 are relatively close to the threshold and significantly smaller than the spike observed at lag 1. For this reason both $q=1$ and $q=4$ are considered and we fit both $ARIMA(0,0,1)(0,1,1)_{12}$ and $ARIMA(0,0,4)(0,1,1)_{12}$. If we consider the PACF there are significant spikes at lags 1 and 3, however as with the ACF the spike at lag 3 is relatively close to the threshold and significantly smaller than the spike observed at lag 1. So we similar consider both $p=1$ and $p=3$ and fit both $ARIMA(1,0,0)(0,1,1)_{12}$ and $ARIMA(3,0,0)(0,1,1)_{12}$. Finally, an ARIMA model is fitted automatically for comparison. 
:::

```{r arima-acf-pacf}
# acf plots on longer lags to see shape 
plt_acf <- train_filled %>% 
              ACF(Temperature, lag_max=120) %>% 
                  autoplot() + 
                  scale_x_continuous(breaks = seq(0, 120, by = 12)) + 
                  theme(axis.title.x = element_blank())
plt_acf_diff <- train_filled %>% 
                    ACF(diff_temp, lag_max=120) %>%
                        autoplot() + 
                        scale_x_continuous(breaks = seq(0, 120, by = 12)) +
                        ylab("acf (Differenced)")

plt_acf + plt_acf_diff + plot_layout(ncol = 1)

# pacf plots on longer lags to see shape 
plt_pacf <- train_filled %>% 
              PACF(Temperature, lag_max=120) %>% 
                  autoplot() + 
                  scale_x_continuous(breaks = seq(0, 120, by = 12)) + 
                  theme(axis.title.x = element_blank())
plt_pacf_diff <- train_filled %>% 
                    PACF(diff_temp, lag_max=120) %>% 
                          autoplot() + 
                          scale_x_continuous(breaks = seq(0, 120, by = 12)) +
                          ylab("pacf (Differenced)")

plt_pacf + plt_pacf_diff + plot_layout(ncol = 1)

# all together on fewer lags
train_filled %>% gg_tsdisplay(diff_temp, plot_type = "partial", lag = 24)
```


```{r arima-fits}
#fit all models
ARIMA_fits <- train_filled %>% model(arima400011 = ARIMA(Temperature ~ pdq(3, 0, 0) + PDQ(0, 1, 1)),
                                     arima100011 = ARIMA(Temperature ~ pdq(1, 0, 0) + PDQ(0, 1, 1)),
                                     arima003011 = ARIMA(Temperature ~ pdq(0, 0, 4) + PDQ(0, 1, 1)), 
                                     arima001011 = ARIMA(Temperature ~ pdq(0, 0, 1) + PDQ(0, 1, 1)), 
                                     auto = ARIMA(Temperature))

```



```{r arima-comparison}
# find best on AICc
report(ARIMA_fits) %>% arrange(AICc) %>% select(.model:BIC)

accuracy(ARIMA_fits) %>% arrange(RMSE)

ARIMA_fits %>% select(arima100011) %>% report() #report best

ARIMA_fits %>% select(auto) %>% report() #report auto
```

------------------------------------------------------------------------

## Selected Model and Justification

::: contributor
Contributors: Bea Cooke
:::

::: discussion
**Selected model:** 

Based on AICc, $ARIMA(1,0,0)(0,1,1)_{12}$ is selected as the model likely to have the best predictive ability. AICc was reported as 2532, this model also reported the lowest uncorrected AIC (2532) and BIC (2547) scores further supporting this choice. 
:::

```{r arima-selected}
#selected model
ARIMA_fit <- train_filled %>% model(arima100011 = ARIMA(Temperature ~ pdq(1, 0, 0) + PDQ(0, 1, 1))) 

#model coefs
ARIMA_fit %>% select(arima100011) %>% coefficients() %>% select(.model:std.error) %>% mutate(across(where(is.numeric), round, 3))
```


------------------------------------------------------------------------

## Model Equation (Backshift Notation)

::: contributor
Contributors: Bea Cooke
:::

::: discussion

$$
(1 - 0.311 B)(1 - B^{12}) y_t = (1 - 0.882 B^{12})\varepsilon_t
$$
:::

------------------------------------------------------------------------

# Dynamic Regression Models

## Explanation of Dynamic Regression

::: contributor
Contributors: Joshua Buchanan
:::

::: discussion
At their core, **dynamic regression models** are what you get when
standard regression and ARIMA are combined: you get external predictors
*and* proper time series error structure in one model.

The three approaches compared:

-   **Standard regression** lets you include useful external variables
    (holidays, economic data etc.) but assumes errors are white noise
-   **Pure ARIMA** handles autocorrelation well but you're stuck with no
    external predictors, just the series modelling itself
-   **Dynamic regression** covers both:
    $y_t = \beta_0 + \beta_1 x_{1,t} + \cdots + \eta_t$ where the
    residual $\eta_t$ follows an $\text{ARIMA}(p,d,q)$ process rather
    than being white noise

One thing worth noting: there are actually **two error terms** in the
model. $\eta_t$ is the regression error, and $\varepsilon_t$ is the
white noise innovation buried inside the ARIMA: only $\varepsilon_t$ is
assumed to be white noise, not $\eta_t$.

The intention is to use dynamic regression when you have predictors that
clearly explain some variation in your series, but your residuals are
still autocorrelated after fitting.
:::

------------------------------------------------------------------------

## Model Exploration

::: contributor
Contributors: Joshua Buchanan
:::

::: discussion
**Approach:**

Three predictor types were considered for the dynamic regression models:

-   **Linear trend** --- included via `trend()` to capture the long-run
    warming signal in the Nelson temperature series
-   **Fourier terms** --- used to model seasonality rather than seasonal
    ARIMA differencing, with $K$ ranging from 1 to 6 (the maximum for
    monthly data with period 12). Each additional $K$ adds one sine and
    one cosine pair, allowing progressively more flexible seasonal
    shapes
-   **Intervention variables** --- visual inspection of the training
    data and filtering for extreme values found no obvious spikes or
    structural breaks, so no dummy variables were included

`PDQ(0,0,0)` was specified to prevent the auto ARIMA from also fitting
seasonal ARIMA components on top of the Fourier terms, which would
double-handle the seasonality. The $p$, $d$, $q$ orders for the error
ARIMA were left to automatic selection in all models.
:::

```{r dynreg-setup}
# Add fourier terms + trend to train_filled
# Try K = 1, 2, 3, 4, 5, 6 (max for period 12)
dynreg_fit <- train_filled %>%
  model(
    K1 = ARIMA(Temperature ~ trend() + fourier(K = 1) + PDQ(0,0,0)),
    K2 = ARIMA(Temperature ~ trend() + fourier(K = 2) + PDQ(0,0,0)),
    K3 = ARIMA(Temperature ~ trend() + fourier(K = 3) + PDQ(0,0,0)),
    K4 = ARIMA(Temperature ~ trend() + fourier(K = 4) + PDQ(0,0,0)),
    K5 = ARIMA(Temperature ~ trend() + fourier(K = 5) + PDQ(0,0,0)),
    K6 = ARIMA(Temperature ~ trend() + fourier(K = 6) + PDQ(0,0,0))
  )
```

```{r dynreg-comparison}
glance(dynreg_fit) %>% arrange(AICc) %>% select(.model, AIC, AICc, BIC)
accuracy(dynreg_fit) %>% select(.model, RMSE, MAE, MASE)
```

```{r dynreg-outlier-check}
# Check training data for obvious outliers/spikes worth modelling
train_filled %>% autoplot(Temperature)

# Identify candidate spike dates if any stand out
train_filled %>% 
  filter(Temperature == max(Temperature) | Temperature == min(Temperature))

# I do not think there are any crazy outliers here - let me know if you disagree. 
```

------------------------------------------------------------------------

## Selected Model and Justification

::: contributor
Contributors: Joshua Buchanan
:::

::: discussion
Based on AICc, **K3** is selected as the best dynamic regression model,
with the lowest AICc of 2435.880 and AIC of 2435.559.

The results show a clear pattern. K1 is by far the worst (AICc
2608.415), K2 improves, and K3 achieves the best fit. Beyond K3, each
additional pair of Fourier terms increases AICc, meaning the added
flexibility is not worth the extra complexity.

K3 includes six Fourier terms (3 sine/cosine pairs), which is sufficient
to capture the dominant seasonal harmonics in monthly temperature data
without overfitting. This aligns with the principle of parsimony: K4--K6
offer no improvement in AICc or BIC, suggesting the higher-frequency
seasonal variation they model is largely noise.

No intervention variables were included. Visual inspection of the
training data and filtering for extreme values revealed no obvious
spikes or structural breaks warranting dummy variables. The series
appears to vary smoothly within a consistent seasonal range throughout
the training period.

**Summary:**

-   K3 achieves the lowest AICc (2435.880) and AIC (2435.559) of all
    candidates
-   BIC also favours K3 (2494.271), confirming the parsimony argument
-   K4--K6 show monotonically increasing AICc, ruling them out
-   K1--K2 underfit the seasonal structure given the sharp AICc drop to
    K3
-   No outlier dummies added: inspection found no extreme values
    justifying intervention terms
:::

```{r dynreg-best-residuals}
# Residual diagnostics on best model (update K after glance above)
# dynreg_fit %>%
#   select(K3) %>%
#   gg_tsresiduals()
# 
# dynreg_fit %>%
#   augment() %>%
#   features(.resid, ljung_box, lag = 24)

## looks good - we should explain further in terms of each plot cand ljung box for completeness. 
# Bea Question: does the dof need to be specified for ljung_box?
```

```{r dynreg-final}
dynreg_final <- train_filled %>%
  model(K3 = ARIMA(Temperature ~ trend() + fourier(K = 3) + PDQ(0,0,0)))
```

------------------------------------------------------------------------

## Trend Interpretation

::: contributor
Contributors: Joshua Buchanan
:::

::: discussion
The trend coefficient is **0.0022°C per month**.

**95% Confidence Interval:**

$$0.0022 \pm 1.96 \times 0.0002 = [0.0018,\ 0.0026]$$

This interval sits entirely above zero, meaning we can confidently
reject the hypothesis that temperatures are staying constant.

**Per-century change:**

$$0.0022 \times 12 \times 100 = \textbf{0.264°C per century}$$

With the 95% CI translating to **[0.216, 0.312]°C per century**.

Average temperatures in Nelson are increasing over the long term, though
the rate is very slow: roughly 0.26°C per century.
:::

```{r dynreg-trend}
report(dynreg_final)
```

### Adding some forecasting for the later section here for this model - just leaving in this section now rather than below as maybe our model will change and easier to edit here before moving later.

```{r dynreg-forecast}
# # Future data for h=12 forecast horizon
# future <- new_data(train_filled, 12)
# 
# dynreg_fc <- dynreg_final %>%
#   forecast(new_data = future)
# 
# # Plot forecast against full nelson series
# dynreg_fc %>% autoplot(nelson)
```

```{r dynreg-accuracy}
# # Compare against test set
# accuracy(dynreg_fc, nelson) %>%
#   select(.model, RMSE, MAE, MAPE, MASE)

```

```{r dynreg-forecast-table}
# # Print point forecasts and 95% intervals
# dynreg_fc %>%
#   hilo(level = 95) %>%
#   select(date, .mean, `95%`)
```

------------------------------------------------------------------------

# Assumption Checking

## Residual Diagnostics

::: contributor
Contributors: [Name(s) here]
:::

```{r residuals-ets}
ETS_final %>% gg_tsresiduals()
```

```{r residuals-arima}
ARIMA_fit %>% gg_tsresiduals()
```

```{r residuals-dynreg}
dynreg_final %>%  gg_tsresiduals()
```

------------------------------------------------------------------------

## Ljung-Box Tests

::: contributor
Contributors: [Name(s) here]
:::

```{r ljung-box}
# --- Ljung-Box tests for all three models ---
ETS_final  %>% augment() %>% features(.innov, ljung_box, lag = 24, dof = 999) # ETS - dof & update model
ARIMA_fit %>% augment() %>% features(.innov, ljung_box, lag = 24, dof = 2) #ARIMA
dynreg_final %>% augment() %>% features(.innov, ljung_box, lag = 24, dof = 11) #DyReg - dof 

```

------------------------------------------------------------------------

## Discussion

::: contributor
Contributors: [Name(s) here]
:::

::: discussion
**Do you trust these models for forecasting?** Discuss the residual
plots and Ljung-Box results for each model. Comment on whether the
assumptions of white noise residuals appear to be met and what
implications this has for forecasting reliability.
:::

------------------------------------------------------------------------

# Forecasting

## Forecasts (h = 12)

::: contributor
Contributors: [Name(s) here]
:::

```{r forecasts-ets}
# --- ETS forecast for h = 12 months ---
ETS_fc_final <- ETS_final %>%
  forecast(h = 12)

plt_ets_fc <- autoplot(ETS_fc_final, nelson)
plt_ets_fc_test <- autoplot(ETS_fc_final, test)

plt_ets_fc + plt_ets_fc_test + plot_layout(ncol = 1)

# Print point forecasts and 95% intervals
ETS_fc_final %>%
  hilo(level = 95) %>%
  select(date, .mean, `95%`)
```

```{r forecasts-arima}
# --- ARIMA forecast for h = 12 months ---
ARIMA_fc <- ARIMA_fit %>% forecast(h = 12)

plt_arima_fc <- autoplot(ARIMA_fc, nelson)
plt_arima_fc_test <-autoplot(ARIMA_fc, test)

plt_arima_fc + plt_arima_fc_test + plot_layout(ncol = 1)

# Print point forecasts and 95% intervals
ARIMA_fc %>%
  hilo(level = 95) %>%
  select(date, .mean, `95%`)
```

```{r forecasts-dynreg}
# --- Dynamic regression forecast for h = 12 months ---
# Future data for h=12 forecast horizon
future <- new_data(train_filled, 12)

dynreg_fc <- dynreg_final %>%
  forecast(new_data = future)

# Plot forecast against full nelson series
plt_dyreg_fc <- dynreg_fc %>% autoplot(nelson)
plt_dyreg_fc_test <- dynreg_fc %>% autoplot(test)

plt_dyreg_fc + plt_dyreg_fc_test + plot_layout(ncol = 1)


# Print point forecasts and 95% intervals
dynreg_fc %>%
  hilo(level = 95) %>%
  select(date, .mean, `95%`)
```

```{r forecasts-plot}
# --- Combined forecast plot comparing all three models ---

```

------------------------------------------------------------------------

## Forecast Error vs Test Set

::: contributor
Contributors: [Name(s) here]
:::

```{r forecast-accuracy}
ETS <- accuracy(ETS_fc_final, nelson) 
AR <- accuracy(ARIMA_fc, nelson)
DR <- accuracy(dynreg_fc, nelson) 

bind_rows(ETS, AR, DR) %>% select(.model, RMSE, MAE, MAPE, MASE)
```

------------------------------------------------------------------------

## Discussion

::: contributor
Contributors: [Name(s) here]
:::

::: discussion
**How well did your models perform?** Compare the forecast accuracy
metrics (e.g. RMSE, MAE, MAPE) across your three models. Discuss which
performed best and offer an explanation for why.
:::

------------------------------------------------------------------------

# LLMs in Time Series Analysis

<!-- Delete whichever option your group did NOT choose -->

## Option A — LLM Point Forecast

::: contributor
Contributors: [Name(s) here]
:::

::: discussion
**Prompt used:**

> *Paste your prompt(s) to the LLM here.*

**LLM's thought process:**

> *Summarise or paste the LLM's reasoning here.*

**LLM point forecast for May 2026:** [value here]
:::

```{r llm-model-forecasts}
# --- Extend your three models to h = 13 to get May 2026 forecasts ---

```

::: discussion
**Comparison and comments:** Compare the LLM's forecast to your three
models' forecasts for May 2026. Comment on anything interesting.
:::

------------------------------------------------------------------------

## Option B — Chronos Model

::: contributor
Contributors: [Name(s) here]
:::

::: discussion
**Setup notes:** Describe how you installed and ran the Chronos model in
Python. Include any relevant details about feeding in the training set.
:::

```{r chronos-results}
# --- Import Chronos forecasts (e.g. from a CSV saved from Python) ---
# --- Compute forecast error on test set ---
# --- Compare to ETS, ARIMA, Dynamic regression ---

```

::: discussion
**Did Chronos perform better or worse?** Compare Chronos forecast
accuracy to your three models. Give a possible explanation for the
result.
:::

------------------------------------------------------------------------

# Summary

::: contributor
Contributors: [Name(s) here]
:::

::: discussion
**Brief overall summary:** Summarise the key findings of your analysis —
which model performed best, what the data showed, and any limitations or
caveats worth noting.
:::

------------------------------------------------------------------------

*Report compiled using R `r R.version$major`.`r R.version$minor` and
fpp3 `r packageVersion("fpp3")`.*
