---
title: "STATS786 Group Project"
author: "Group 4: Joshua Buchanan, Bea Cooke, Xinyi Kong, Marlon Chhour"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    toc_depth: 3
    number_sections: true
    df_print: paged
    code_folding: show
---

```{css, echo=FALSE}
.contributor {
  background-color: #2c3e50;
  color: #ecf0f1;
  padding: 4px 10px;
  border-radius: 4px;
  font-size: 0.8em;
  font-family: monospace;
  display: inline-block;
  margin-bottom: 12px;
}

.discussion {
  background-color: #f8f9fa;
  border-left: 4px solid #18bc9c;
  padding: 12px 16px;
  margin: 12px 0;
  border-radius: 0 4px 4px 0;
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo      = TRUE,
  warning   = FALSE,
  message   = FALSE,
  fig.align = "center",
  fig.width = 9,
  fig.height = 5
)
```

```{r libraries, message=FALSE, warning=FALSE}
library(tidyverse)
library(fpp3)
library(tsibble)
library(feasts)
library(fable)
library(lubridate)
library(here)
library(urca)
```

---

# Data Cleaning and Exploratory Data Analysis

## Data Description

<div class="contributor">Contributors: Joshua Buchanan</div>

::: {.discussion}
The dataset contains 993 observations of monthly average temperature (°C) recorded at a single location, spanning from April 1943 to approximately mid-2025: a period of over 80 years. Each row represents one month, with the month name and year stored as separate columns. 
:::

```{r load-data}
# Load data
nelson <- read_csv('nelson.csv') 
glimpse(nelson)
```


---

## Train / Test Split

<div class="contributor">Contributors: Joshua Buchanan</div>

::: {.discussion}
The training set excludes the most recent 12 months, which form the test set. All model fitting is performed on the training set only.
:::

```{r train-test-split}
# Turn into tsibble first
nelson <- nelson %>%
  mutate(date = yearmonth(paste(Month, Year))) %>%
  select(date, Temperature) %>%
  as_tsibble(index = date)
glimpse(nelson)

# Create training and test sets: test is last 12 months.
train <- nelson %>% filter(date < max(date) - 11)
test <- nelson %>% filter(date >= max(date) - 11)

# Checking number of rows
nrow(train)
nrow(test)

# Checking the range too. 
range(train$date)
range(test$date)
```

---

## Missing Values

<div class="contributor">Contributors: Joshua Buchanan</div>

::: {.discussion}
We have four missing values. What should we do? *Maybe we should talk about this as a team. *
:::

```{r missing-values}
has_gaps(nelson)  # check for gaps in time index
gaps <- scan_gaps(nelson) # shows where the gaps are
gaps 

# There are no gaps in the Test set so we only need to update the training set. 

## interpolation with ARIMA as per fpp3 sec:13.9
nel_miss <- train %>% fill_gaps()
train_filled <- nel_miss %>% 
                model(ARIMA(Temperature)) %>% 
                interpolate(nel_miss)

has_gaps(train_filled) 
train_filled %>% filter(date %in% gaps$date)
```

```{r, fig.cap="Time plot of Nelson temperatures with interpolated temperature values for missing dates."}
train_filled %>% autoplot(Temperature)
```

```{r, fig.cap="Subseries time plot of Nelson temperatures until 1947, temperature value interpolated at March 1946, denoted as a red vertical line."}
train_filled %>% 
  filter(date < yearmonth("1947 Jan")) %>% 
  autoplot(Temperature) + 
    geom_vline(xintercept = yearmonth("1946 Mar"), 
               colour = "red")
```

---

## Visualisation and Feature Discussion

<div class="contributor">Contributors: Joshua Buchanan</div>

::: {.discussion}
The time series displays strong, consistent seasonality with temperatures peaking in January (summer, ~17–18°C) and reaching a trough in July (winter, ~6–7°C). The seasonal subseries plot confirms this pattern is highly stable across all 80 years, with monthly means well-defined and little variation in the seasonal shape over time. The STL decomposition confirms the seasonality is additive, with the seasonal component oscillating consistently between approximately −6°C and +3°C throughout the entire series, showing no evidence of changing amplitude over time.

The trend component from the STL decomposition reveals a positive trend, rising from approximately 11°C in the early 1940s to around 14°C by the mid-2020s. The trend is not monotonically increasing, with a notable dip around the 1970s–1980s before accelerating upward from the 1990s onward. The remainder component appears to behave like white noise, fluctuating randomly within approximately ±2°C with no obvious structure, large outliers, or evidence of heteroscedasticity, which is encouraging for modelling. 

The ACF confirms strong positive autocorrelation at lags of 12, 24, and 36 months and strong negative autocorrelation at lags of 6, 18, and 30 months, consistent with the annual seasonal cycle, while the PACF shows significant spikes across the first 12 lags suggesting a complex short-term autocorrelation structure that will need to be accounted for in both the ARIMA and dynamic regression models.
:::

```{r time-plot}
# Time plot
train_filled %>% autoplot(Temperature)
```

```{r subseries-plot}
train_filled %>% gg_subseries()
```

```{r}
# Seasonality plot 
train_filled %>% gg_season(Temperature)
```

```{r}
# ACF plot
train_filled %>% ACF(Temperature, lag_max=48) %>% autoplot()
```

```{r}
# PACF plot
train_filled %>% PACF(Temperature, lag.max=48) %>% autoplot()
```

```{r}
# STL decomp
train_filled %>%
  model(STL(Temperature ~ trend(window=21))) %>%
  components() %>%
  autoplot()
```

---

# ETS Models

## Model Exploration and Shortlisting

<div class="contributor">Contributors: Xinyi Kong</div>

::: {.discussion}
**Methodology:
Candidate ETS models were selected based on strong annual seasonality, a slowly evolving trend, and approximately constant variance over time. These features suggest that additive error and additive seasonal structures are appropriate for this dataset.
The manually selected ETS models include:
## ETS model equations

### ETS(AAA):
\[
y_t = \ell_{t-1} + b_{t-1} + s_{t-m} + \epsilon_t
\]

### ETS(ANA):
\[
y_t = \ell_{t-1} + s_{t-m} + \epsilon_t
\]

### ETS(AAdA):
\[
y_t = \ell_{t-1} + \phi b_{t-1} + s_{t-m} + \epsilon_t
\]

### Multiplicative benchmark models:

### ETS(M,A,A):
\[
y_t = (\ell_{t-1} + b_{t-1} + s_{t-m})(1 + \epsilon_t)
\]

### ETS(M,Ad,A):
\[
y_t = (\ell_{t-1} + \phi b_{t-1} + s_{t-m})(1 + \epsilon_t)
\]
Given these features, additive error and additive seasonal structures were considered the primary modelling choice. The initial shortlist focused on ETS models with additive seasonality, including:

ETS(AAA): full additive specification with trend and seasonality
ETS(ANA): no-trend specification to test whether a trend component is necessary
ETS(AAdA): damped trend specification to improve long-term forecasting stability

In addition, multiplicative specifications were included as benchmark models to assess whether a proportional error and seasonal structure could provide improved fit. These include:

ETS(M,A,A)
ETS(M,Ad,M)

Although multiplicative models are typically suited to series with variance increasing with the level of the series, they were included for completeness and empirical comparison.

An automatic ETS procedure was also fitted to select an optimal model based on AICc across the full ETS model space.

An automatic ETS model was also fitted using ETS(Temperature), which selects the optimal combination of error, trend, and seasonal components by minimising AICc over all possible ETS specifications.
:::

```{r ets-manual}
# Manually fitted ETS models
library(imputeTS)

ETS_fit <- train_filled %>%
  fill_gaps() %>%
  mutate(Temperature = na_seadec(Temperature)) %>%
  model(additive = ETS(Temperature ~ error("A") + trend("A") + season("A")),
        ANA = ETS(Temperature ~ error("A") + trend("N") + season("A")),
        AAdA = ETS(Temperature ~ error("A") + trend("Ad") + season("A")),
        multiplicative = ETS(Temperature ~ error("M") + trend("A") + season("M")),
        hw = ETS(Temperature ~ error("M") + trend("Ad") + season("M")))

ETS_fc = ETS_fit %>% 
  forecast(h = 12) 

ETS_fc %>%
  autoplot(nelson)
```

```{r ets-auto}
# Automatically fitted ETS model
auto_fit = train_filled %>%
  fill_gaps() %>%
  mutate(Temperature = na_seadec(Temperature)) %>%
  model(ETS(Temperature))
report(auto_fit) 

## AAdA becomes the best now here too with the NAs filled via ARIMA in the training data.
```

```{r ets-comparison}
# Model comparison (e.g. AICc, BIC, accuracy)
accuracy(ETS_fit) %>%
  select(.model, RMSE, MAE, MAPE, MASE)

glance(ETS_fit) %>% arrange(AICc)
library(feasts)

ETS_fit %>%
  select(additive) %>%
  gg_tsresiduals()

ETS_fit %>%
  select(ANA) %>%
  gg_tsresiduals()

ETS_fit %>%
  select(AAdA) %>%
  gg_tsresiduals()

ETS_fit %>%
  augment() %>%
  features(.resid, ljung_box, lag = 24)
```

---

## Selected Model and Justification

<div class="contributor">Contributors: Xinyi Kong</div>

::: {.discussion}
**Selected model:**
[Note from Joshua - now the AAdA is outperforming the ANA model in terms of AIC and AICc since we have switched to using the training set with the filled values - what do you think of this?]

Based on AICc and forecast accuracy, the ETS(ANA) model is selected as the preferred model. It achieves the lowest AICc and competitive error measures, while maintaining a simpler structure without a trend component.

Although Ljung-Box tests indicate remaining autocorrelation in the residuals for all candidate models, this is common in long seasonal time series such as temperature data and does not invalidate the model for forecasting purposes.

Therefore, ETS(ANA) is selected as the final model due to its balance of parsimony, interpretability, and predictive performance.
:::

```{r ets-selected}
# Selected ETS model
ETS_final <- train %>%
  fill_gaps() %>%
  mutate(Temperature = na_seadec(Temperature)) %>%
  model(ANA = ETS(Temperature ~ error("A") + trend("N") + season("A")))
report(ETS_final)

ETS_fc_final <- ETS_final %>%
  forecast(h = 12)
autoplot(ETS_fc_final, train)
```

---

## Model Equations

<div class="contributor">Contributors: Xinyi Kong</div>

::: {.discussion}

- **Measurement equation:**
\[
y_t = \ell_{t-1} + s_{t-m} + \epsilon_t
\]

- **State equations:**

Level:
\[
\ell_t = \ell_{t-1} + \alpha \epsilon_t
\]

Seasonal:
\[
s_t = s_{t-m} + \gamma \epsilon_t
\]

- **Parameter estimates:**

\[
\alpha = 0.1076, \quad \gamma = 0.0001008, \quad \sigma^2 = 0.7298
\]

[NOTE: we will have to change the above since top model has changed?]
:::

---

# ARIMA Models

## Transformations and Differencing

<div class="contributor">Contributors: [Name(s) here]</div>

::: {.discussion}
**Transformations:** Describe any Box-Cox or log transformations applied and why.

**Differencing:** Describe the differencing required to achieve stationarity — use ACF/PACF and unit root tests to justify.
:::

```{r arima-transformations}
# --- Transformation and differencing exploration ---

```

```{r arima-acf-pacf}
# --- ACF and PACF plots ---

```

```{r arima-unit-root}
# --- Unit root tests (e.g. KPSS, ADF) ---

```

---

## Model Exploration and Shortlisting

<div class="contributor">Contributors: [Name(s) here]</div>

::: {.discussion}
**Methodology:** Describe how you built your shortlist of ARIMA models from the ACF/PACF and automatic selection.
:::

```{r arima-manual}
# --- Manually specified ARIMA models ---

```

```{r arima-auto}
# --- Automatically fitted ARIMA model (e.g. ARIMA(y)) ---

```

```{r arima-comparison}
# --- Model comparison ---

```

---

## Selected Model and Justification

<div class="contributor">Contributors: [Name(s) here]</div>

::: {.discussion}
**Selected model:** State your chosen ARIMA model and justify your selection.
:::

```{r arima-selected}
# --- Fit your selected ARIMA model here ---

```

---

## Model Equation (Backshift Notation)

<div class="contributor">Contributors: [Name(s) here]</div>

::: {.discussion}
**Write down the fitted model equation in backshift notation here.**

$$
(1 - \phi_1 B)(1 - B)^d y_t = (1 + \theta_1 B)\varepsilon_t
$$

*Replace with your actual equation and parameter estimates.*
:::

---

# Dynamic Regression Models

## Explanation of Dynamic Regression

<div class="contributor">Contributors: Joshua Buchanan</div>

::: {.discussion}
At their core, **dynamic regression models** are what you get when standard regression and ARIMA
are combined: you get external predictors *and* proper time series error structure in one model.

The three approaches compared:

- **Standard regression** lets you include useful external variables (holidays, economic data etc.)
but assumes errors are white noise
- **Pure ARIMA** handles autocorrelation well but you're stuck with no external predictors, just
the series modelling itself
- **Dynamic regression** covers both: $y_t = \beta_0 + \beta_1 x_{1,t} + \cdots + \eta_t$ where
the residual $\eta_t$ follows an $\text{ARIMA}(p,d,q)$ process rather than being white noise

One thing worth noting: there are actually **two error terms** in the model. $\eta_t$ is the
regression error, and $\varepsilon_t$ is the white noise innovation buried inside the ARIMA: only
$\varepsilon_t$ is assumed to be white noise, not $\eta_t$.

The intention is to use dynamic regression when you have predictors that clearly explain some variation in your series,
but your residuals are still autocorrelated after fitting.
:::

---

## Model Exploration

<div class="contributor">Contributors: Joshua Buchanan</div>

::: {.discussion}
**Approach:** Describe the predictors you included — linear trend, Fourier terms for seasonality, and any intervention variables (e.g. outlier spikes). Explain your choices.
:::

```{r dynreg-setup}
# Dynamic regression models with Fourier terms for seasonality
library(fpp3)

# Add fourier terms + trend to train_filled
# Try K = 1, 2, 3, 4, 5, 6 (max for period 12)
dynreg_fit <- train_filled %>%
  model(
    K1 = ARIMA(Temperature ~ trend() + fourier(K = 1) + PDQ(0,0,0)),
    K2 = ARIMA(Temperature ~ trend() + fourier(K = 2) + PDQ(0,0,0)),
    K3 = ARIMA(Temperature ~ trend() + fourier(K = 3) + PDQ(0,0,0)),
    K4 = ARIMA(Temperature ~ trend() + fourier(K = 4) + PDQ(0,0,0)),
    K5 = ARIMA(Temperature ~ trend() + fourier(K = 5) + PDQ(0,0,0)),
    K6 = ARIMA(Temperature ~ trend() + fourier(K = 6) + PDQ(0,0,0))
  )
```


```{r dynreg-comparison}
glance(dynreg_fit) %>% arrange(AICc) %>% select(.model, AIC, AICc, BIC)
accuracy(dynreg_fit) %>% select(.model, RMSE, MAE, MASE)
```

```{r dynreg-outlier-check}
# Check training data for obvious outliers/spikes worth modelling
train_filled %>% autoplot(Temperature)

# Identify candidate spike dates if any stand out
train_filled %>% 
  filter(Temperature == max(Temperature) | Temperature == min(Temperature))

# I do not think there are any crazy outliers here - let me know if you disagree. 
```





---

## Selected Model and Justification

<div class="contributor">Contributors: Joshua Buchanan</div>

::: {.discussion}
Based on AICc, **K3** is selected as the best dynamic regression model, with the lowest AICc of
2435.880 and AIC of 2435.559.

The results show a clear pattern. K1 is by far the worst (AICc 2608.415), K2 improves, and K3 achieves the best fit. 
Beyond K3, each additional pair of Fourier terms
increases AICc, meaning the added flexibility is not worth the extra complexity.

K3 includes six Fourier terms (3 sine/cosine pairs), which is sufficient to capture the dominant
seasonal harmonics in monthly temperature data without overfitting. This aligns with the principle
of parsimony: K4--K6 offer no improvement in AICc or BIC, suggesting the higher-frequency seasonal
variation they model is largely noise.

No intervention variables were included. Visual inspection of the training data and filtering for
extreme values revealed no obvious spikes or structural breaks warranting dummy variables. The
series appears to vary smoothly within a consistent seasonal range throughout the training period.

**Summary:**

- K3 achieves the lowest AICc (2435.880) and AIC (2435.559) of all candidates
- BIC also favours K3 (2494.271), confirming the parsimony argument
- K4--K6 show monotonically increasing AICc, ruling them out
- K1--K2 underfit the seasonal structure given the sharp AICc drop to K3
- No outlier dummies added: inspection found no extreme values justifying intervention terms
:::

```{r dynreg-best-residuals}
# Residual diagnostics on best model (update K after glance above)
dynreg_fit %>%
  select(K3) %>%
  gg_tsresiduals()

dynreg_fit %>%
  augment() %>%
  features(.resid, ljung_box, lag = 24)

## looks good - we should explain further in terms of each plot cand ljung box for completeness. 
```

```{r dynreg-final}
dynreg_final <- train_filled %>%
  model(K3 = ARIMA(Temperature ~ trend() + fourier(K = 3) + PDQ(0,0,0)))
```

---

## Trend Interpretation

<div class="contributor">Contributors: Joshua Buchanan</div>

::: {.discussion}
The trend coefficient is **0.0022°C per month**.

**95% Confidence Interval:**

$$0.0022 \pm 1.96 \times 0.0002 = [0.0018,\ 0.0026]$$

This interval sits entirely above zero, meaning we can confidently reject the hypothesis that
temperatures are staying constant.

**Per-century change:**

$$0.0022 \times 12 \times 100 = \textbf{0.264°C per century}$$

With the 95% CI translating to **[0.216, 0.312]°C per century**.

Average temperatures in Nelson are increasing over the long term, though the rate is very slow:
roughly 0.26°C per century. 
:::

```{r dynreg-trend}
report(dynreg_final)
```



### Adding some forecasting for the later section here for this model - just leaving in this section now rather than below as maybe our model will change and easier to edit here before moving later. 


```{r dynreg-forecast}
# Future data for h=12 forecast horizon
future <- new_data(train_filled, 12)

dynreg_fc <- dynreg_final %>%
  forecast(new_data = future)

# Plot forecast against full nelson series
dynreg_fc %>% autoplot(nelson)
```

```{r dynreg-accuracy}
# Compare against test set
accuracy(dynreg_fc, nelson) %>%
  select(.model, RMSE, MAE, MAPE, MASE)

```

```{r dynreg-forecast-table}
# Print point forecasts and 95% intervals
dynreg_fc %>%
  hilo(level = 95) %>%
  select(date, .mean, `95%`)
```


---

# Assumption Checking

## Residual Diagnostics

<div class="contributor">Contributors: [Name(s) here]</div>

```{r residuals-ets}
# --- Residual diagnostic plots for ETS model ---

```

```{r residuals-arima}
# --- Residual diagnostic plots for ARIMA model ---

```

```{r residuals-dynreg}
# --- Residual diagnostic plots for Dynamic Regression model ---

```

---

## Ljung-Box Tests

<div class="contributor">Contributors: [Name(s) here]</div>

```{r ljung-box}
# --- Ljung-Box tests for all three models ---

```

---

## Discussion

<div class="contributor">Contributors: [Name(s) here]</div>

::: {.discussion}
**Do you trust these models for forecasting?** Discuss the residual plots and Ljung-Box results for each model. Comment on whether the assumptions of white noise residuals appear to be met and what implications this has for forecasting reliability.
:::

---

# Forecasting

## Forecasts (h = 12)

<div class="contributor">Contributors: [Name(s) here]</div>

```{r forecasts-ets}
# --- ETS forecast for h = 12 months ---

```

```{r forecasts-arima}
# --- ARIMA forecast for h = 12 months ---

```

```{r forecasts-dynreg}
# --- Dynamic regression forecast for h = 12 months ---

```

```{r forecasts-plot}
# --- Combined forecast plot comparing all three models ---

```

---

## Forecast Error vs Test Set

<div class="contributor">Contributors: [Name(s) here]</div>

```{r forecast-accuracy}
# --- Compute forecast error against test set for all three models ---
# --- e.g. accuracy(forecasts, test) ---

```

---

## Discussion

<div class="contributor">Contributors: [Name(s) here]</div>

::: {.discussion}
**How well did your models perform?** Compare the forecast accuracy metrics (e.g. RMSE, MAE, MAPE) across your three models. Discuss which performed best and offer an explanation for why.
:::

---

# LLMs in Time Series Analysis

<!-- Delete whichever option your group did NOT choose -->

## Option A — LLM Point Forecast

<div class="contributor">Contributors: [Name(s) here]</div>

::: {.discussion}
**Prompt used:**

> *Paste your prompt(s) to the LLM here.*

**LLM's thought process:**

> *Summarise or paste the LLM's reasoning here.*

**LLM point forecast for May 2026:** [value here]
:::

```{r llm-model-forecasts}
# --- Extend your three models to h = 13 to get May 2026 forecasts ---

```

::: {.discussion}
**Comparison and comments:** Compare the LLM's forecast to your three models' forecasts for May 2026. Comment on anything interesting.
:::

---

## Option B — Chronos Model

<div class="contributor">Contributors: [Name(s) here]</div>

::: {.discussion}
**Setup notes:** Describe how you installed and ran the Chronos model in Python. Include any relevant details about feeding in the training set.
:::

```{r chronos-results}
# --- Import Chronos forecasts (e.g. from a CSV saved from Python) ---
# --- Compute forecast error on test set ---
# --- Compare to ETS, ARIMA, Dynamic regression ---

```

::: {.discussion}
**Did Chronos perform better or worse?** Compare Chronos forecast accuracy to your three models. Give a possible explanation for the result.
:::

---

# Summary

<div class="contributor">Contributors: [Name(s) here]</div>

::: {.discussion}
**Brief overall summary:** Summarise the key findings of your analysis — which model performed best, what the data showed, and any limitations or caveats worth noting.
:::

---

*Report compiled using R `r R.version$major`.`r R.version$minor` and fpp3 `r packageVersion("fpp3")`.*
